---
title: "Statistical assignment 4"
author: "[Daniel Orchard]"
date: "[26 Feb 2020]"
output: github_document
---

```{r setup, include=FALSE}
# Please note these options.
# This tells R Markdown that we want to show code in the output document.
knitr::opts_chunk$set(echo = TRUE)
# Switching off messages in the output document. 
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
# Switching on caching to make things faster (don't commit cache files on Github).
knitr::opts_chunk$set(cache = TRUE)
```

In this assignment you will need to reproduce 5 ggplot graphs. I supply graphs as images; you need to write the ggplot2 code to reproduce them and knit and submit a Markdown document with the reproduced graphs (as well as your .Rmd file).

First we will need to open and recode the data. I supply the code for this; you only need to change the file paths.

    ```{r}
library(tidyverse)
Data8 <- read_tsv("C:/Users/dan19/Documents/MY DOCUMENTS/Year 2 Term 2 Modules/Data III/Understanding Society Data/UKDA-6614-tab/tab/ukhls_w8/h_indresp.tab")
Data8 <- Data8 %>%
        select(pidp, h_age_dv, h_payn_dv, h_gor_dv)
Stable <- read_tsv("C:/Users/dan19/Documents/MY DOCUMENTS/Year 2 Term 2 Modules/Data III/datan3_2020/data/UKDA-6614-tab/tab/ukhls_wx/xwavedat.tab")
Stable <- Stable %>%
        select(pidp, sex_dv, ukborn, plbornc)
Data <- Data8 %>% left_join(Stable, "pidp")
rm(Data8, Stable)
Data <- Data %>%
        mutate(sex_dv = ifelse(sex_dv == 1, "male",
                           ifelse(sex_dv == 2, "female", NA))) %>%
        mutate(h_payn_dv = ifelse(h_payn_dv < 0, NA, h_payn_dv)) %>%
        mutate(h_gor_dv = recode(h_gor_dv,
                         `-9` = NA_character_,
                         `1` = "North East",
                         `2` = "North West",
                         `3` = "Yorkshire",
                         `4` = "East Midlands",
                         `5` = "West Midlands",
                         `6` = "East of England",
                         `7` = "London",
                         `8` = "South East",
                         `9` = "South West",
                         `10` = "Wales",
                         `11` = "Scotland",
                         `12` = "Northern Ireland")) %>%
        mutate(placeBorn = case_when(
                ukborn  == -9 ~ NA_character_,
                ukborn < 5 ~ "UK",
                plbornc == 5 ~ "Ireland",
                plbornc == 18 ~ "India",
                plbornc == 19 ~ "Pakistan",
                plbornc == 20 ~ "Bangladesh",
                plbornc == 10 ~ "Poland",
                plbornc == 27 ~ "Jamaica",
                plbornc == 24 ~ "Nigeria",
                TRUE ~ "other")
        )
    ```

Reproduce the following graphs as close as you can. For each graph, write two sentences (not more!) describing its main message.  

# 1. Univariate distribution (20 points).

    ```{r}
 
    ggplot(Data, mapping = aes(x = h_payn_dv)) +
    geom_freqpoly() +
    xlab("Net Monthly Pay") +
    ylab("Number of Respondents")
    
    
    ```
    
The message of this graph is that net monthly pay is clustered around the ~1300 pounds mark. After that it tails off quite fast with a small bump again at 5500 pounds.

# 2. Line chart (20 points). The lines show the non-parametric association between age and monthly earnings for men and women.

    ```{r}
    
    ggplot(Data, mapping = aes(x = h_age_dv, y = h_payn_dv,linetype = sex_dv)) +
    geom_smooth(color = "black") +
    xlab("Age") +
    ylab("Monthly Earnings") +
    xlim(16,65) 
    
   
    
    
    
    ```

This chart shows that monthly earnings rise with age, peaking at about 45 before declining. It also shows that men on average earn more than women at every age point, but particluary from ages 25 onwards, where the gap starts to widen. 

# 3. Faceted bar chart (20 points).

    ```{r}
    

    
bysex <- Data %>%
  group_by(sex_dv, placeBorn) %>%
  summarise(
    medianwage = median(h_payn_dv, na.rm = TRUE)) %>% 
    filter(!is.na(sex_dv)) %>%
    filter(!is.na(placeBorn))

 bysex %>%   
    ggplot(aes(x = sex_dv, y = medianwage)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ placeBorn, ncol = 3) +
    ylim(0,2000) +
    xlab("Sex") +
    ylab("Median Monthly Net Pay")

    
    ```

These charts show that regardless of place of birth that men on average have a higher median monthly net pay. The largest differnece in pay parity appears to be for UK born whilst the lowest is with Bangledeshi families.


# 4. Heat map (20 points).

    ```{r}
    
byregion <- Data %>%
  group_by(h_gor_dv, placeBorn) %>%
  summarise(
    meanage = mean(h_age_dv, na.rm = TRUE)) %>% 
    filter(!is.na(h_gor_dv)) %>%
    filter(!is.na(placeBorn)) 
 
 byregion %>%  ggplot( aes(x = h_gor_dv, y = placeBorn, fill = meanage)) +
    geom_tile() +
    xlab("Region")+
    ylab("Country of Birth") +
    theme(axis.text.x = element_text(angle = 90))
        
    
    
    ```

This heat map is designed to show the mean age of residents of each region of the UK depending on their Country of birth. The chart shows that the average age seems to be around 60-70 but there are some much lower mean ages such as for Nigerian born people in Scotland, however this may be due to a low sample there. There are also a few NAs, particulary in Northern Ireland.

# 5. Population pyramid (20 points).

    ```{r}
    
 #group by age and then by sex
    # then do a count of each age group by sex. then should be about to do it
    
population <- Data %>%  group_by(sex_dv,h_age_dv) %>% 
    count(h_age_dv, sex_dv)   
    
population$n <- ifelse(population$sex_dv == "male", -1*population$n, population$n)
    
 ggplot(population, aes(x = h_age_dv, y = n, fill = sex_dv)) + 
  geom_bar(data = subset(population, sex_dv == "female"), stat = "identity",      colour = "red") +
  geom_bar(data = subset(population, sex_dv == "male"), stat = "identity", colour = "blue") + 
  coord_flip() +
  xlab("Age")
    
    
    ```
    
This population pyramid shows that the number of people in age group starts quite high at 0-25, decreases a bit (possibly due to emmigration), then reexpands for 30-65 before starting to shrink over 75, presumably as people start to die. It also shows that for most age groups there are more women than men, especially in old age.